Changes in accessibility to emergency and community food services during COVID-19 and implications for low income populations in Hamilton, Ontario

### 2021-08-27 Accessibility-Foodbanks-Hamilton/Response-to-Reviewers-R1.Rmd

Reviewer \#1: Overall, I found the paper well-written and clear. I did have some concerns, primarily with the paper's framing and the conclusions drawn:

\textcolor{blue}{Thank you for the opportunity to revise the paper, and for the valuable suggestions to do so.}

1.  The characterization of the emergency food services included is a little unclear. It seems until the table on page 10 that the focus is on food banks, but really it looks like the focus is on ALL emergency food provision (including communal meal services). The framing of the piece is therefore a little "off", as meal programs have a much longer history (pre-Victorian) than food banks, and are intended to support a somewhat different clientele (e.g., those without kitchen/cooking facilities, and usually "singles" rather than families). I think it would make sense for the authors to use the broader term "emergency food services" in the title and throughout, and to change the framing at the beginning (please note that I don't think it is necessary to go into the history of emergency food provision here at all really... but the current presentation beginning with food banks is misleading).

\textcolor{blue}{Thank you for the comment. We agree with the Reviewer’s critique regarding the framing of the paper and have adopted the suggestion to reframe the facilities as “emergency and community food services” as they do serve different roles. In addition to the terminology, we have also added a paragraph describing the role of food banks as well as other food services in reducing food insecurity in Section 2.}

@@

2.  Relatedly, the conflation of food banks and meal programs is a bit of a challenge in terms of interpreting the data. For example, one would EXPECT congregate dining to essentially shut down during the pandemic, as this could be a route for disease spread. In some cases, some kind of food services may have continued to be offered (e.g., through takeout), but in general closures would be expected. In the food BANKS, however, the mode of delivery might need to be modified slightly, but direct impacts of covid would be less significant. Additionally, as mentioned above these services target different demographics. So, by conflating these two modes of food delivery, the utility of the analysis is compromised.

\textcolor{blue}{By considering all facilities equally, we agree that the analysis does conflate the different service types offered. We better acknowledge this through the clarified terminology of "emergency and community food services" recommended above. However, all of these service types play an important role in food (in)security in the study area and our overarching focus remains on geographic accessibility and adaptation in the food access landscape pre- and during-COVID in Hamilton. Beyond the role of the transportation network, a particular interest is in how floating catchment methods can offer insight into how the closure of some of these important services affects demand and “congestion” at those that remain open.}

@@

3.  The presentation of data in relation to food security is a bit confusing, as data from pre-covid, during covid, and from both the U.S. and Canada is presented. This makes it a little challenging to draw out the contemporary situation in the study area, and should be presented more clearly (and given that they seem to have food security data from May 2020, and there is widely available data from 2018 as well, they SHOULD be able to give a clearer sense of how much food security has increased during the pandemic in Canada). Also, I think comparisons to the situation in the U.S. during covid should be made cautiously, as their social interventions have been very different than Canada's (and in both countries, responses have varied widely be state/province).

\textcolor{blue}{We agree with the critique and have streamlined the paper's introduction and Section 2 to offer greater clarity on the changing situation in Canada and Hamilton and reduce the comparisons with the United States.}

@@

4.  Overall, the paper seems longer than it needs to be in the introduction/context and conclusion sections - there is quite a bit of background that seems unnecessary, and unnecessary restatement of that material in the conclusion.

\textcolor{blue}{In response to this comment, we have streamlined both the introduction and concluding sections to remove restated and inapplicable information (e.g. the US context information in the introduction). Further changes to the conclusion in response to critique 5 are noted below.}

@@

Specific comments:

Pg 2 paragraph 2 - a little unclear which of the data presented is pre-covid and which is during covid

\textcolor{blue}{This paragraph has been updated to clarify the pre-covid timeline of some of the papers.}

@@

Pg 2 ln 46 (and also pg 21 ln 33) - were 40% of actual visitors children, or are they extrapolating from households with children? It seems unlikely that parents are actually bringing their children along, so "visitors" may be the wrong word ("users"?)

\textcolor{blue}{•	This text has been updated to reflect the Reviewer’s suggestion regarding “users” rather than visitors.}

@@

Pg 4 paragraph 1 - not sure there is much evidence that connects food insecurity (which is usually "mild" in relation to true hunger/severe food\
insecurity) in developed nations with nutrient deficiency? If so, this should be shared. If not, the argument should be clarified in relation to known outcomes (e.g., stress, obesity)

\textcolor{blue}{Thank you for this comment. We have re-examined the cited literature and the Kirkpatrick and Tarasuk article “Food Insecurity Is Associated with Nutrient Inadequacies among Canadian Adults and Adolescents’ in the Journal of Nutrition from 2008 is the most direct in this regard. This paper is more clearly cited and additional references noting the poorer physical and mental health outcomes associated with food insecurity have been added to clarify the evidence.}

@@

Pg 4 paragraph 2 - not sure that food banks include meal programs/prepared meals (the food BANK wouldn't, although the organization might offer both)

\textcolor{blue}{This section has been revised to focus first on food insecurity and food bank use for emergency food access. Furthermore, a paragraph has been added to discuss the role of community food services in reducing food insecurity, such as community kitchens and congregate dining. We believe this addition strengthens the paper’s link to food insecurity and thank Reviewer 1 for raising this critique.}

@@

Pg 5 ln 27 - seems to imply that the U.S. experience of increasing food insecurity during the pandemic is true in Canada, but the U.S. context in terms of emergency income relief is very different, and the May 2020 statistic of 14.7% is not contextualised (is this an increase? From what?)

\textcolor{blue}{Additional text is now included to note that the 14.7% in May 2020 is up from 10.5% reported in 2017-2018. Furthermore, the text about the US and Canada is now more clearly delineated.}

@@

Pg 23 (last page) - would be nice to have a final concluding sentence that restates the contribution of the paper, rather than ending with limitations.

\textcolor{blue}{In addition to the changes noted above, the concluding section has been re-organized to be more impactful with regards to its conclusions.}

### 2021-08-27 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.Rmd
cdh R1 edits

# load updated files manually
load("output-data-files/ttm_transit.RData")
load("./output-data-files/foodbank_table.RData")
```

@@

Economic stress caused by the COVID-19 pandemic, rising unemployment rates, and changes in poverty levels have disrupted the food environment and led to higher rates of food insecurity [@niles2020early]. Pandemic-related food security studies in the US have found a substantial increase in households experiencing food insecurity for the first time, and also in households experiencing more severe food insecurity than before [@niles2020early; @wolfson2020food]. Most recently in May 2020, Canada recorded 14.7% of its population living in food insecurity in the past 30 days, up from 10.5% in 2017-2018 [@statisticscanada2020food]. Considering the negative mental and physical health effects, increases in food insecurity rates due to the onset of COVID-19, signal a change in the food environment with potential damages to population health during the course of the pandemic and beyond [@niles2020early]. Recent data from FBC [@fbc2020] showed that 52% of member food banks reported an increase in usage in March of 2020 when initial lockdown restrictions were put in place across much of the country. The pandemic also created significant staffing issues with 42% of food banks reporting a reduction in volunteers. However, 53% of food banks later reported a decrease in use into the summer of 2020 which FBC members attributed to emergency financial support programs from the federal government. Nevertheless, some of these benefit programs were temporary. Although more recent statistics on food bank use in Hamilton in 2020 and 2021 are not yet available, data from the Daily Bread Food Bank [@dbfb2020] in neighbouring Toronto for August 2020 shows visits climbing 51% year-over-year in that city, which suggests that many households in Hamilton are likely to turn to food bank services to meet their needs.

@@

Food insecurity is a significant issue for many Canadian households and is associated with reductions in nutrition and poor mental and physical health outcomes. While emergency and community food services can provide some relief, the levels of need in the City of Hamilton were on the rise before the onset of the COVID-19 pandemic, with a troubling increase in the number of visits made by children to these services. While temporary financial relief measures may have helped in the summer of 2020, the pandemic has in all probability increased food insecurity for many households. To compound matters, the pandemic has also resulted in major disruptions, including to employment, mobility alternatives, and to emergency and community food services. In response, this research has sought to better understand accessibility to food banks and food service locations, as well as how the closure of some locations over the pre- and during-COVID time periods affected the potential for low-income households to reach these amenities. Using the balanced floating catchment area method to account for population demand and congestion effects at service points, we estimated multi-modal accessibility to emergency and community food service locations for low-income households. The weighting of travel time estimates by the modal split in different zonal geographies tailors the results to patterns of travel behaviour captured in the regional travel survey. Moreover, our parcel-level analysis presents a disaggregate approach to estimating accessibility based on the locations of residential parcels and dwellings.

Previous work has noted the important role of geography alongside other socio-economic and demographic indicators in household access to healthy food. The present papers is, to the best of our knowledge, among the first studies to focus on the geographic component of accessibility to emergency food services [@allen2021Changes]. Beyond accounting for the inflation of demand that occurs in traditional floating catchment area methods, our application of the balanced floating catchment area approach offers a novel analysis of accessibility to emergency and communal food services that is sensitive to the locations of low-income households, information on their age distributions and typical trip-making behaviour, the locations of services, their operations over time, and the characteristics of the city's multi-modal transportation network. Our results show that while accessibility levels were lower in the city's more car-oriented suburban and rural areas to begin with, the closure of 14% of the city's emergency and community food service locations during the pandemic resulted in an overall decrease in accessibility across the city. However, these effects were not uniform over space or for different population groups. Since the balanced floating catchment area method takes into account changes in demand and congestion for service providers, the closure of some services reverberates throughout the catchment areas of the whole city. For some suburban zones, the closure of a relatively high level-of-service location results in the remaining services being spread over a larger population. In others, high auto dependence for trips leads to decreases in accessibility that accumulate due to the loss of several locations initially reachable within 15 minutes by car. Reductions in accessibility in the city's more urban north end, where low-income households conduct higher proportions of trips by transit and walking, emphasize the importance of geographic proximity in the potential to reach service locations for these residents. Beyond geography, the results also highlight the differential impact of closures during COVID-19 on population groups, with seniors and children beign the two most impacted groups.

In the absence of information regarding how food insecure households travel to food banks and related services, we examined accessibility to food banks using a 15 minutes (one-way) travel time threshold. The fact that we must rely on a standard created for accessible drinking water in the developing world only serves to highlight the tragedy of food insecurity in an affluent country like Canada. More broadly, it points to the absurd need to understand how a bad situation was made worse by the pandemic: in effect, the analysis reveals that disparities in the need for emergency and community food services predated the pandemic, that the pandemic contributed to the deterioration of these services, and that populations already in distress, particularly children, ended up in an even more adverse state. How much worse, it is impossible to say, mainly because there is also a dearth of information, let alone standards, regarding _acceptable_ or _sufficient_ level of service when it comes to emergency food services. In fact, we would argue that the standard should be that no household faced food insecurity. As others have noted [e.g., @poppendieck1999sweet; @men2021food] the root of food insecurity is income poverty. There is likely a place for emergency food and community food services. In addition to providing food, these services could satisfy social needs by offering a social setting for seniors or by helping to connect households in need with longer term supports. From a food security perspective, on the other hand, these services should work only as a short term solution, and not as a semi-permanent feature of life for some of our fellow human beings. From a human rights perspective, long-term reliance on emergency food services should be as unacceptable in Canada as lack of clean drinking water within 30 minutes is elsewhere. Thus, while our analysis is valuable to map the suffering caused by food insecurity, from a policy perspective maintaining a robust social safety net that includes Employment Insurance and paid sick days are better tools to reduce this suffering than increasing the accessibility of emergency food services for food insecure populations.

In terms of future research, it is important to note that the degree to which our low-income cut-off of CAD\$40,000 reflects food insecurity in the different zones of the study area is not known. We also consider all emergency and communal food service locations equally. Information on differential capacities at food provider locations is currently not collected, and given the closure of facilities was not possible to obtain. Data on services, such as the number of meals served, could be used to refine the analysis in the future. Moreover, while the travel survey allows us to model multi-modal accessibilities that align with travel behaviour observed in the travel survey and capture differences in accessibility by age categories, the use of travel survey data for modelling food insecurity also has its limitations. Research into the population weighting methods used in the TTS note that the survey may under-count the lowest- and highest-income households in the survey study region, although the magnitude of this under-counting is unknown, and approximately 20% of respondents to the 2016 survey did not report their income [@rose2018transportation]. In that regard, the modal splits of low-income households observed through the TTS data may not accurately reflect the travel behaviours of food insecure populations, and our estimates might in fact be somewhat conservative if those households who do not report income rely more on walking and/or transit for their mobility needs.

>>>>

# load updated files manually
load("./output-data-files/ttm_transit.RData")
load("./output-data-files/foodbank_table.RData")
```

@@

In the United States, economic stress caused by the COVID-19 pandemic, rising unemployment rates, and changes in poverty levels have disrupted the food environment and led to higher rates of food insecurity [@niles2020early]. Pandemic-related food security studies in the US have found a substantial increase in households experiencing food insecurity for the first time, and also in households experiencing more severe food insecurity than before [@niles2020early; @wolfson2020food]. Considering the negative mental and physical health effects, increases in food insecurity rates due to the onset of COVID-19, signal a change in the food environment with potential damages to population health during the course of the pandemic and beyond [@niles2020early].

Canada has also seen an increase in the number of households living in food insecurity due to the COVID-19 pandemic. Survey results from Statistics Canada from May of 2020 suggest that 14.7% of the population was living in food insecurity in the past 30 days, up from 10.5% in 2017-2018 [@statisticscanada2020food]. Recent data from FBC [@fbc2020] showed that 52% of member food banks reported an increase in usage in March of 2020 when initial lockdown restrictions were put in place across much of the country. The pandemic also created significant staffing issues with 42% of food banks reporting a reduction in volunteers. However, 53% of food banks later reported a decrease in use into the summer of 2020 which FBC members attributed to emergency financial support programs from the federal government. Nevertheless, some of these benefit programs were temporary. Although more recent statistics on food bank use in Hamilton in 2020 and 2021 are not yet available, data from the Daily Bread Food Bank [@dbfb2020] in neighbouring Toronto for August 2020 shows visits climbing 51% year-over-year in that city, which suggests that many households in Hamilton are likely to turn to food bank services to meet their needs.

@@

Food insecurity is a significant issue for many Canadian households and while emergency and community food services can provide some relief, the COVID-19 pandemic has in all probability increased food insecurity for many households. To compound matters, the pandemic has also resulted in major disruptions, including to employment, mobility alternatives, and to emergency and community food services. In response, this research has sought to better understand accessibility to food banks and food service locations, as well as how the closure of some locations over the pre- and during-COVID time periods affected the potential for low-income households to reach these amenities.

Previous work has noted the important role of geography alongside other socio-economic and demographic indicators in household access to healthy food. The present papers is, to the best of our knowledge, among the first studies to focus on the geographic component of accessibility to emergency food services [@allen2021Changes]. Using the balanced floating catchment area method to account for population demand and congestion effects at service points, we estimated multi-modal accessibility to emergency and community food service locations for low-income households. The weighting of travel time estimates by the modal split in different zonal geographies tailors the results to patterns of travel behaviour captured in the regional travel survey. Moreover, our parcel-level analysis presents a disaggregate approach to estimating accessibility based on the locations of residential parcels and dwellings. Beyond accounting for the inflation of demand that occurs in traditional floating catchment area methods, our application of the balanced floating catchment area approach offers a novel analysis of accessibility to emergency and communal food services that is sensitive to the locations of low-income households, information on their age distributions and typical trip-making behaviour, the locations of services, their operations over time, and the characteristics of the city's multi-modal transportation network.

Our results show that while accessibility levels were lower in the city's more car-oriented suburban and rural areas to begin with, the closure of 14% of the city's emergency and community food service locations during the pandemic resulted in an overall decrease in accessibility across the city. However, these effects were not uniform over space or for different population groups. Since the balanced floating catchment area method takes into account changes in demand and congestion for service providers, the closure of some services reverberates throughout the catchment areas of the whole city. For some suburban zones, the closure of a relatively high level-of-service location results in the remaining services being spread over a larger population. In others, high auto dependence for trips leads to decreases in accessibility that accumulate due to the loss of several locations initially reachable within 15 minutes by car. Reductions in accessibility in the city's more urban north end, where low-income households conduct higher proportions of trips by transit and walking, emphasize the importance of geographic proximity in the potential to reach service locations for these residents. Beyond geography, the results also highlight the differential impact of closures during COVID-19 on population groups, with seniors and children being the two most impacted groups.

It is important to note that the degree to which our low-income cut-off of CAD\$40,000 reflects food insecurity in the different zones of the study area is not known. We also consider all emergency and communal food service locations equally. Information on differential capacities at food provider locations is currently not collected, and given the closure of facilities was not possible to obtain. Data on services, such as the number of meals served, could be used to refine the analysis in the future. Moreover, while the travel survey allows us to model multi-modal accessibilities that align with travel behaviour observed in the travel survey and capture differences in accessibility by age categories, the use of travel survey data for modelling food insecurity also has its limitations. Research into the population weighting methods used in the TTS note that the survey may under-count the lowest- and highest-income households in the survey study region, although the magnitude of this under-counting is unknown, and approximately 20% of respondents to the 2016 survey did not report their income [@rose2018transportation]. In that regard, the modal splits of low-income households observed through the TTS data may not accurately reflect the travel behaviours of food insecure populations, and our estimates might in fact be somewhat conservative if those households who do not report income rely more on walking and/or transit for their mobility needs.

In the absence of information regarding how food insecure households travel to food banks and related services, we examined accessibility to food banks using a 15 minutes (one-way) travel time threshold. The fact that we must rely on a standard created for accessible drinking water in the developing world only serves to highlight the tragedy of food insecurity in an affluent country like Canada. More broadly, it points to the absurd need to understand how a bad situation was made worse by the pandemic: in effect, the analysis reveals that disparities in the need for emergency and community food services predated the pandemic, that the pandemic contributed to the deterioration of these services, and that populations already in distress, particularly children, ended up in an even more adverse state. How much worse, it is impossible to say, mainly because there is also a dearth of information, let alone standards, regarding _acceptable_ or _sufficient_ level of service when it comes to emergency food services.

While on the one hand this work suggests that inequities in the accessibility to emergency and community food services could be improved through accessibility standards that promote changes in the geographic distribution of service locations and transportation network characteristics, in fact, we would argue that the standard should be that no household faced food insecurity. As others have noted [e.g., @poppendieck1999sweet; @men2021food] the root of food insecurity is income poverty and unless it is eliminated, there will continue to be a place for emergency food and community food services. In addition to providing food, these services satisfy social needs by offering a social setting for seniors or by helping to connect households in need with longer term supports. From a food security perspective, on the other hand, these services should work only as a short term solution, and not as a semi-permanent feature of life for some of our fellow human beings. From a human rights perspective, long-term reliance on emergency food services should be as unacceptable in Canada as lack of clean drinking water within 30 minutes is elsewhere. Thus, while our analysis is valuable to map the suffering caused by food insecurity, from a policy perspective, maintaining a robust social safety net that includes Employment Insurance and paid sick days are better tools to reduce this suffering than increasing the accessibility of emergency food services for food insecure populations.

### 2021-08-27 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.tex
cdh R1 edits

Economic stress caused by the COVID-19 pandemic, rising unemployment
rates, and changes in poverty levels have disrupted the food environment
and led to higher rates of food insecurity (Niles et al., 2020).
Pandemic-related food security studies in the US have found a
substantial increase in households experiencing food insecurity for the
first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Most recently in May 2020, Canada recorded 14.7\% of its population
living in food insecurity in the past 30 days, up from 10.5\% in
2017-2018 (Statistics Canada, 2020a). Considering the negative mental
and physical health effects, increases in food insecurity rates due to
the onset of COVID-19, signal a change in the food environment with
potential damages to population health during the course of the pandemic
and beyond (Niles et al., 2020). Recent data from FBC (FBC, 2020) showed
that 52\% of member food banks reported an increase in usage in March of
2020 when initial lockdown restrictions were put in place across much of
the country. The pandemic also created significant staffing issues with
42\% of food banks reporting a reduction in volunteers. However, 53\% of
food banks later reported a decrease in use into the summer of 2020
which FBC members attributed to emergency financial support programs
from the federal government. Nevertheless, some of these benefit
programs were temporary. Although more recent statistics on food bank
use in Hamilton in 2020 and 2021 are not yet available, data from the
Daily Bread Food Bank (DBFB, 2020) in neighbouring Toronto for August
2020 shows visits climbing 51\% year-over-year in that city, which
suggests that many households in Hamilton are likely to turn to food
bank services to meet their needs.

@@

Food insecurity is a significant issue for many Canadian households and
is associated with reductions in nutrition and poor mental and physical
health outcomes. While food bank use can provide some relief, the levels
of need in the City of Hamilton were on the rise before the onset of the
COVID-19 pandemic, with a troubling increase in the number of visits
made by children. While temporary financial relief measures may have
helped reduce food bank use in the summer of 2020, the pandemic has in
all probability increased food insecurity for many households. To
compound matters, the pandemic has also resulted in major disruptions,
including to employment, mobility alternatives, and to food bank and
food service delivery. In response, this research has sought to better
understand accessibility to food banks and food service locations, as
well as how the closure of some locations over the pre- and during-COVID
time periods affected the potential for low-income households to reach
these amenities. Using the balanced floating catchment area method to
account for population demand and congestion effects at service points,
we estimated multi-modal accessibility to food bank and food service
locations for low-income households. The weighting of travel time
estimates by the modal split in different zonal geographies tailors the
results to patterns of travel behaviour captured in the regional travel
survey. Moreover, our parcel-level analysis presents a disaggregate
approach to estimating accessibility based on the locations of
residential parcels and dwellings.

Our results show that while accessibility levels were lower in the
city's more car-oriented suburban and rural areas to begin with, the
closure of 14\% of the city's food bank and low-cost or free food
service locations during the pandemic resulted in an overall decrease in
accessibility across the city. However, these effects were not uniform
over space or for different population groups. Since the balanced
floating catchment area method takes into account changes in demand and
congestion for service providers, the closure of food bank and food
service locations can reverberate throughout the catchment areas of the
whole city. For some suburban zones, the closure of a relatively high
level-of-service location results in the remaining services being spread
over a larger population. In others, high auto dependence for trips
leads to decreases in accessibility that accumulate due to the loss of
several food bank and service locations initially reachable within 15
minutes by car. Reductions in accessibility in the city's more urban
north end, where low-income households conduct higher proportions of
trips by transit and walking, emphasize the importance of geographic
proximity in the potential to reach food bank and service locations for
these residents. Beyond geography, the results also highlight the
differential impact of food banks closing during COVID-19 on population
groups, with seniors and children most affected.

In the absence of information regarding how food insecure households
travel to food banks and related services, we examined accessibility to
food banks using a 15 minutes (one-way) travel time threshold. The fact
that we must rely on a standard created for accessible drinking water in
the developing world only serves to highlight the tragedy of food
insecurity in an affluent country like Canada. More broadly, it points
to the absurd need to understand how a bad situation was made worse by
the pandemic: in effect, the analysis reveals that disparities in
accessibility to food banks and services predated the pandemic, that the
pandemic contributed to the deterioration of these services, and that
populations already in distress, particularly children, ended up in an
even more adverse state. How much worse, it is impossible to say, mainly
because there is also a dearth of information, let alone standards,
regarding what is an \emph{acceptable} or \emph{sufficient} level of
service when it comes to food banks. Whether this is one facility per
1,000 population or three facilities per 1,000 population, or any other
number, we know not. But like international human rights standards for
accessibility to clean drinking water within 30 minutes (round-trip),
this work strongly suggests that adopting better defined standards for
access to food banks that move beyond generalized conceptualizations of
``access'' to consider the geographic component of accessibility (and
its interactions with travel behaviour and individual/household
socioeconomic and demographic traits) would make a meaningful
contribution to understanding food access and ``food bank deserts'' for
food insecure populations.

In summary, while previous work has noted the important role of
geography alongside other socio-economic and demographic indicators in
household access to healthy food, this is, to the best of our knowledge,
the first study that has focused on the geographic component of food
bank accessibility. Beyond accounting for the inflation of demand that
occurs in traditional floating catchment area methods, our application
of the balanced floating catchment area approach offers a novel analysis
of food bank and food service accessibility that is sensitive to the
locations of low-income households, information on their age
distributions and typical trip-making behaviour, food bank and food
service locations and their operations over time, and the
characteristics of the city's multi-modal transportation network.
Despite the limitations noted above, the findings from this analysis can
provide quantitative evidence to help urban planners and community
administrators strategically allocate food bank services considering
accessibility for different population groups, especially when
considering the unequal distribution of impacts from the COVID-19
pandemic.

In terms of future research, it is important to note that the degree to
which our low-income cut-off of CAD\$40,000 reflects food insecurity in
the different zones of the study area is not known. We also consider all
food banks and food service locations equally. Information on
differential capacities at food provider locations is currently not
collected, and given the closure of facilities was not possible to
obtain. Data on services, such as the number of meals served, could be
used to refine the analysis in the future. Moreover, while the travel
survey allows us to model multi-modal accessibilities that align with
travel behaviour observed in the travel survey and capture differences
in accessibility by age categories, the use of travel survey data for
modelling food insecurity also has its limitations. Research into the
population weighting methods used in the TTS note that the survey may
under-count the lowest- and highest-income households in the survey
study region, although the magnitude of this under-counting is unknown,
and approximately 20\% of respondents to the 2016 survey did not report
their income (Rose, 2018). In that regard, the modal splits of
low-income households observed through the TTS data may not accurately
reflect the travel behaviours of food insecure populations, and our
estimates might in fact be somewhat conservative if those households who
do not report income rely more on walking and/or transit for their
mobility needs.

@@

@@

>>>>

In the United States, economic stress caused by the COVID-19 pandemic,
rising unemployment rates, and changes in poverty levels have disrupted
the food environment and led to higher rates of food insecurity (Niles
et al., 2020). Pandemic-related food security studies in the US have
found a substantial increase in households experiencing food insecurity
for the first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Considering the negative mental and physical health effects, increases
in food insecurity rates due to the onset of COVID-19, signal a change
in the food environment with potential damages to population health
during the course of the pandemic and beyond (Niles et al., 2020).

Canada has also seen an increase in the number of households living in
food insecurity due to the COVID-19 pandemic. Survey results from
Statistics Canada from May of 2020 suggest that 14.7\% of the population
was living in food insecurity in the past 30 days, up from 10.5\% in
2017-2018 (Statistics Canada, 2020a). Recent data from FBC (FBC, 2020)
showed that 52\% of member food banks reported an increase in usage in
March of 2020 when initial lockdown restrictions were put in place
across much of the country. The pandemic also created significant
staffing issues with 42\% of food banks reporting a reduction in
volunteers. However, 53\% of food banks later reported a decrease in use
into the summer of 2020 which FBC members attributed to emergency
financial support programs from the federal government. Nevertheless,
some of these benefit programs were temporary. Although more recent
statistics on food bank use in Hamilton in 2020 and 2021 are not yet
available, data from the Daily Bread Food Bank (DBFB, 2020) in
neighbouring Toronto for August 2020 shows visits climbing 51\%
year-over-year in that city, which suggests that many households in
Hamilton are likely to turn to food bank services to meet their needs.

@@

Food insecurity is a significant issue for many Canadian households and
while emergency and community food services can provide some relief, the
COVID-19 pandemic has in all probability increased food insecurity for
many households. To compound matters, the pandemic has also resulted in
major disruptions, including to employment, mobility alternatives, and
to emergency and community food services. In response, this research has
sought to better understand accessibility to food banks and food service
locations, as well as how the closure of some locations over the pre-
and during-COVID time periods affected the potential for low-income
households to reach these amenities.

Previous work has noted the important role of geography alongside other
socio-economic and demographic indicators in household access to healthy
food. The present papers is, to the best of our knowledge, among the
first studies to focus on the geographic component of accessibility to
emergency food services (Allen and Farber, 2021). Using the balanced
floating catchment area method to account for population demand and
congestion effects at service points, we estimated multi-modal
accessibility to emergency and community food service locations for
low-income households. The weighting of travel time estimates by the
modal split in different zonal geographies tailors the results to
patterns of travel behaviour captured in the regional travel survey.
Moreover, our parcel-level analysis presents a disaggregate approach to
estimating accessibility based on the locations of residential parcels
and dwellings. Beyond accounting for the inflation of demand that occurs
in traditional floating catchment area methods, our application of the
balanced floating catchment area approach offers a novel analysis of
accessibility to emergency and communal food services that is sensitive
to the locations of low-income households, information on their age
distributions and typical trip-making behaviour, the locations of
services, their operations over time, and the characteristics of the
city's multi-modal transportation network.

Our results show that while accessibility levels were lower in the
city's more car-oriented suburban and rural areas to begin with, the
closure of 14\% of the city's emergency and community food service
locations during the pandemic resulted in an overall decrease in
accessibility across the city. However, these effects were not uniform
over space or for different population groups. Since the balanced
floating catchment area method takes into account changes in demand and
congestion for service providers, the closure of some services
reverberates throughout the catchment areas of the whole city. For some
suburban zones, the closure of a relatively high level-of-service
location results in the remaining services being spread over a larger
population. In others, high auto dependence for trips leads to decreases
in accessibility that accumulate due to the loss of several locations
initially reachable within 15 minutes by car. Reductions in
accessibility in the city's more urban north end, where low-income
households conduct higher proportions of trips by transit and walking,
emphasize the importance of geographic proximity in the potential to
reach service locations for these residents. Beyond geography, the
results also highlight the differential impact of closures during
COVID-19 on population groups, with seniors and children being the two
most impacted groups.

It is important to note that the degree to which our low-income cut-off
of CAD\$40,000 reflects food insecurity in the different zones of the
study area is not known. We also consider all emergency and communal
food service locations equally. Information on differential capacities
at food provider locations is currently not collected, and given the
closure of facilities was not possible to obtain. Data on services, such
as the number of meals served, could be used to refine the analysis in
the future. Moreover, while the travel survey allows us to model
multi-modal accessibilities that align with travel behaviour observed in
the travel survey and capture differences in accessibility by age
categories, the use of travel survey data for modelling food insecurity
also has its limitations. Research into the population weighting methods
used in the TTS note that the survey may under-count the lowest- and
highest-income households in the survey study region, although the
magnitude of this under-counting is unknown, and approximately 20\% of
respondents to the 2016 survey did not report their income (Rose, 2018).
In that regard, the modal splits of low-income households observed
through the TTS data may not accurately reflect the travel behaviours of
food insecure populations, and our estimates might in fact be somewhat
conservative if those households who do not report income rely more on
walking and/or transit for their mobility needs.

In the absence of information regarding how food insecure households
travel to food banks and related services, we examined accessibility to
food banks using a 15 minutes (one-way) travel time threshold. The fact
that we must rely on a standard created for accessible drinking water in
the developing world only serves to highlight the tragedy of food
insecurity in an affluent country like Canada. More broadly, it points
to the absurd need to understand how a bad situation was made worse by
the pandemic: in effect, the analysis reveals that disparities in the
need for emergency and community food services predated the pandemic,
that the pandemic contributed to the deterioration of these services,
and that populations already in distress, particularly children, ended
up in an even more adverse state. How much worse, it is impossible to
say, mainly because there is also a dearth of information, let alone
standards, regarding \emph{acceptable} or \emph{sufficient} level of
service when it comes to emergency food services.

While on the one hand this work suggests that inequities in the
accessibility to emergency and community food services could be improved
through accessibility standards that promote changes in the geographic
distribution of service locations and transportation network
characteristics, in fact, we would argue that the standard should be
that no household faced food insecurity. As others have noted (Men and
Tarasuk, 2021; e.g., Poppendieck, 1999) the root of food insecurity is
income poverty and unless it is eliminated, there will continue to be a
place for emergency food and community food services. In addition to
providing food, these services satisfy social needs by offering a social
setting for seniors or by helping to connect households in need with
longer term supports. From a food security perspective, on the other
hand, these services should work only as a short term solution, and not
as a semi-permanent feature of life for some of our fellow human beings.
From a human rights perspective, long-term reliance on emergency food
services should be as unacceptable in Canada as lack of clean drinking
water within 30 minutes is elsewhere. Thus, while our analysis is
valuable to map the suffering caused by food insecurity, from a policy
perspective, maintaining a robust social safety net that includes
Employment Insurance and paid sick days are better tools to reduce this
suffering than increasing the accessibility of emergency food services
for food insecure populations.

@@

\leavevmode\hypertarget{ref-allen2021Changes}{}%
Allen, J., Farber, S., 2021. Changes in transit accessibility to food
banks in toronto during COVID-19. Findings.
doi:\href{https://doi.org/10.32866/001c.24072}{10.32866/001c.24072}

@@

\leavevmode\hypertarget{ref-poppendieck1999sweet}{}%
Poppendieck, J., 1999. Sweet charity?: Emergency food and the end of
entitlement. Penguin.
